Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2024-10-16 manuscript.Rmd
more revision updates

@@

An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change. Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs. Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (`r format_range(range(cc$crab_data_shallow_unscaled$depth))` m) across the ~2,500 km span of the Hawaiian Archipelago. We compare these assemblages with those sampled along a depth gradient (`r format_range(range(cc$crab_data$depth[cc$crab_data$island == "Oahu"]))` m) on O&#x02BB;ahu. We recovered `r num(sum(sample_sums(cc$crabs_untransformed)),1)` individuals representing `r sum(unlist(all_summary$types_breakdown))` taxa from `r all_summary$taxonomy_breakdown$families` families. Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth. However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.

@@

Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [(cite?)]{.mark}. We targeted brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011], recovered from these structures to examine drivers of community composition across the Hawaiian Archipelago. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.

@@

A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (\< 30 m) coral reef habitat (`r format_range(cc$crabs_shallow_untransformed %>% sample_tibble() %>% pull(depth))` m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]). Only units deployed on outer reefs were included in this study. To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016]. Except when discussing specific depth zones, we will refer to any units deployed \< 30 m as "shallow" and those deployed $\geq$ 30 m as "deep".

@@

Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020]. Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:env_table; @tbl:arms_metadata_table]). Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets. The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019]. The index of potential larval immigration was imported from [@selkoeDNACoralReef2016], and represents the potential of a given island to act as a destination for larval settlement. It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp. mat.]. For details regarding data import, preparation, and synthesis, see Supplemental material. Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).

@@

# Discussion {-}
Marine communities are structured by a suite of factors, but there remains no consensus for either generalized mechanisms or the relative importance of drivers of community structure for coral reefs. Our best knowledge of the role of depth in structuring marine communities comes from the intertidal, and points to biotic factors (e.g., predation, competition) determining lower limits and abiotic factors (e.g., desiccation, thermal stress) setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974]. There are numerous examples from other study systems in which the relative contributions of biotic versus abiotic factors to structuring communities run in opposing directions across some habitat gradient [e.g., climate, soil type, elevation, etc. @bradyEvolutionaryEcologyPlant2005; @moore10PlantCompetition2011; @wishingradTemperateZoneIsolation2023]. However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013]. In the Hawaiian archipelago, we found significant differences in the composition of coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI. Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.

The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space. Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016]. Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016]. The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity. Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016]. Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago. What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.

Even though our shallow-water ARMS units were deployed within a narrow depth range (8--17 m), depth still emerged as a significant and consistent driver of community variation among these units. Depth in and of itself is unlikely to affect marine community composition. At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022]. Unfortunately, we lack environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed. Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago. As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.

While depth was a significant driver of community composition among shallow-water ARMS units, its influence was even more apparent when comparing shallow to mesophotic samples. Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]). According to our results, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence. The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples. These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019]. The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community. This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to represent an upper-to-lower transition zone).

In addition to community composition, mesophotic samples also exhibited alpha diversity differences. Effective Simpson diversity was significantly (if slightly) higher in deep communities (for comparisons both among all islands and solely within the MHI) ([@fig:fig_alpha; @fig:fig_alpha_main]). This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although it may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).

Our depth transect was limited to O&#x02BB;ahu, but community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago. Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023]. In that study, communities were found to have more affinity across sites within depth zones than among depth zones within sites. These results and the patterns we observed suggest that depth-associated factors exert stronger ecological influence than those that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.

Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found. However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019]. Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]). Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed. In contrast, the lower depth boundaries of mesophotic reef communities are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure.

Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago. The logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative. Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a]. Crab communities are unlikely to be directly affected by light variability. However, habitat-building species such as corals, algae, and sponges may [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them. If patterns of community structure are driven primarily from the bottom up by light, that may explain our observation of a significant depth effect over less than 10 m in our shallow samples. The importance of light could be tested by sampling communities from the same irradiance level but different depths due to factors altering light penetration to determine if we see consistent patterns of community zonation. In contrast, other factors (such as temperature, dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our comparison of the MHI & NWHI. Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation. Regardless of the exact environmental and ecological drivers at play, it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal. Given the rapid changes in important factors like light, temperature, and nutrients (among others) throughout the upper ~100--150 m of the water column, the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth, whereas *upper* limits may be driven by ecological interactions.

By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of the significant drivers of community structure. Our analysis corroborates previous work and shows depth is the strongest driver of community structure. We find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS. We conclude that some factor(s) that covary with depth exert more influence on the cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.

@@

```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in a CAP ordination by depth zone', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
dists <- bind_rows(
  tibble(dist=shallow_dist,depth="Shallow"),
  tibble(dist=deep_dist,depth="Deep")
) %>%
  mutate(depth = factor(depth,levels=c("Shallow","Deep")))

>>>>

# quickie to just give us a depth range
dr <- function(ps) {
  ps %>%
    sample_tibble() %>%
    pull(depth) %>%
    format_range()
}

@@

An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change. Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs. Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (`r dr(cc$crabs_shallow_untransformed)` m) across the ~2,500 km span of the Hawaiian Archipelago. We compare these assemblages with those sampled along a depth gradient (`r format_range(range(cc$crab_data$depth[cc$crab_data$island == "Oahu"]))` m) on O&#x02BB;ahu. We recovered `r num(sum(sample_sums(cc$crabs_untransformed)),1)` individuals representing `r sum(unlist(all_summary$types_breakdown))` taxa from `r all_summary$taxonomy_breakdown$families` families. Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth. However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.

@@

Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017]. We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011]. We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago, compared crab communities among islands and regions (MHI vs. NWHI), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.

@@

A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reef habitat in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]). Only units deployed on outer reefs were included in this study. To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:arms_table]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016]. Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".

@@

Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020]. Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:env_table; @tbl:arms_metadata_table]). Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets. The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019]. The index of potential larval immigration was imported from [@selkoeDNACoralReef2016] and represents the potential of a given island to act as a destination for larval settlement. It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp. mat.]. For details regarding data import, preparation, and synthesis, see Supplemental material. Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).

@@

# Discussion {-}
The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space. Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016]. Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016]. The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity. Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016]. Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago. What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.

Even though our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units. Depth in and of itself is unlikely to affect marine community composition. At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022]. Unfortunately, we lack environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed. Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago. As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.

While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda]), its influence was even more striking when comparing shallow to mesophotic samples. Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]). According to these analyses, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence. The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples. These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019]. The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community. This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to encompass a transition zone).

In addition to community composition, mesophotic samples also exhibited alpha diversity differences. Effective Simpson diversity was significantly higher in deep communities (for comparisons both among all islands and solely within the MHI) ([@fig:fig_alpha; @fig:fig_alpha_main]). This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although the differences we found were slight and our findings may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).

Our depth transect was limited to O&#x02BB;ahu, but community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago. Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023]. In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites. These results and the patterns we observed suggest that depth-associated factors exert stronger ecological influence than those that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.

Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found. However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019]. Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]). Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed---although due to its significant role as a driver of shallow community variation, it cannot be ruled out. In contrast, the lower depth boundaries of mesophotic ecosystems are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure across depths. Crab communities are unlikely to be directly affected by light variability. However, habitat-building species such as corals, algae, and sponges may be [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991]. If patterns of community structure are driven primarily from the bottom up by light, that may contribute to our observation of a significant depth effect over less than 10 m in our shallow samples.

Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago. Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a]. If mesophotic communities are found to be different from their shallow counterparts across multiple islands (ideally comparing sites experiencing similar irradiance), this suggests that they are vertically structured by a factor (e.g., light) outside of the numerous natural gradients that exist across the Hawaiian Islands. In contrast, other factors (such as temperature, dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our spatial comparison of the MHI & NWHI. Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation. The logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.

Marine communities are structured by a suite of factors, but there remains no consensus for either generalized mechanisms or the relative importance of drivers of community structure for coral reefs. Our best knowledge of the role of depth in structuring marine communities comes from the intertidal, and points to biotic factors (e.g., predation, competition) determining lower limits and abiotic factors (e.g., desiccation, thermal stress) setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974]. There are numerous examples from other study systems in which the relative contributions of biotic versus abiotic factors to structuring communities run in opposing directions across some habitat gradient [e.g., climate, soil type, elevation, etc. @bradyEvolutionaryEcologyPlant2005; @moore10PlantCompetition2011; @wishingradTemperateZoneIsolation2023]. However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013]. In the Hawaiian archipelago, we found significant differences in the composition of shallow coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI. Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors. Despite the significance of these factors in driving shallow community variation across the archipelago, mesophotic assemblages were found to be distinct ([@fig:fig_cap]). Although there were variables we did not consider in our model, this suggests that vertical distributions are influenced by factors other than those driving spatial variation aross the island chain. In light of this, and since mesophotic communities are generally less biodiverse than shallow reefs [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009; @pyleComprehensiveInvestigationMesophotic2016], it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal. Given the rapid changes in important factors like temperature, nutrients, and light throughout the upper ~100--150 m of the water column (particularly the diminishment of photosynthetically active radiation below 100 m), the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth. Conversely, the *upper* limits of the mesophotic zone (the transition from shallow to mesophotic habitat), which occur at the higher end of the depth-biodiversity gradient [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009] may be driven by ecological interactions [@vellendEffectsDiversityDiversity2008].

By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of the significant drivers of community structure. We additionally find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS. Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure. We conclude that some factor(s) covarying with depth exert more influence on cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.

@@

```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in CAP ordination by depth zone', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
dists <- bind_rows(
  tibble(dist=shallow_dist,depth="Shallow"),
  tibble(dist=deep_dist,depth="Deep")
) %>%
  mutate(depth = factor(depth,levels=c("Shallow","Deep")))
